---
title: "Running Around R with Marathons"
author: "Jessica Alexander"
output:
  html_document: default
  pdf_document: default
---

### On your marks, get set, go!

```{r include=FALSE}
df <- read.csv("world_marathon_majors.csv")
library(ggplot2)
library(gganimate)
```

Starting with a quick exploration to see what we're working with. Here are the first six rows of our dataset:
```{r echo=FALSE}
head(df)
```

And here are details of the data available in each column:
```{r echo=FALSE}
str(df)
```

(Under the hood, I'm doing a quick check for missing data.  Good news, no missing values!)
```{r include=FALSE}
check_na <- function(df){
  isna <- c()
  for(i in 1:ncol(df)){
    isna <- c(isna, sum(is.na(df[,i])))
  }
isna
}
check_na(df)
```

First, then, let's fix the `$gender` and `$marathon` columns to make them into factors, and check out the details of each.
```{r echo=FALSE}
df$gender <- as.factor(df$gender)
summary(df$gender)
df$marathon <- as.factor(df$marathon)
summary(df$marathon)
```

Next, let's convert winning times from hh:mm:ss to total seconds to make it more usable.
```{r message=FALSE, echo=FALSE}
library(lubridate)
time_data <- df$time
temp <- hms(time_data)
new_times <- as.integer(hour(temp)*3600 + minute(temp)*60 + second(temp))
df$time <- new_times
summary(df$time)
```
Is that `12600` accurate? [Sure is](https://en.wikipedia.org/wiki/List_of_winners_of_the_Boston_Marathon#Women's_open_division)!

Ok, what about our overall time frame?
```{r echo=FALSE}
summary(df$year)
```

Hmm... We might want to plot chronology, but the original dataset doesn't include which month each race typically takes place, so let's add that. Both Boston and London typically take place in April, but Boston is usually first, so we'll make London "4.5" to Boston's "4".
```{r echo=FALSE}
month <- c()
for(i in 1:nrow(df)){
  if(df$marathon[i]=="Berlin"){
    month <- c(month, 9.0)
  } else if(df$marathon[i]=="Boston"){
    month <- c(month, 4.0)
  } else if(df$marathon[i]=="Chicago"){
    month <- c(month, 10.0)
  } else if(df$marathon[i]=="London"){
    month <- c(month, 4.5)
  } else if(df$marathon[i]=="NYC"){
    month <- c(month, 11.0)
  }else {
    month <- c(month, 2.0)
  }
}
df$month <- as.factor(month)
name <- c("Feb", "Apr", "Apr", "Sep", "Oct", "Nov")
count <- summary(df$month)
display_months <- data.frame(rbind(name, count))
colnames(display_months) <- c("2", "4", "4.5", "9", "10", "11")
display_months
```

We should probably tease apart the men's results from the women's, so let's create two separate dataframes and re-order them chronologically.

#### **Women**
```{r echo=FALSE}
women_only <- df$gender == "Female"
womens_races <- df[women_only,]
womens_races <- womens_races[order(womens_races[,1], womens_races[,7]),]
summary(womens_races)
```
#### **Men**
```{r echo=FALSE}
men_only <- df$gender == "Male"
mens_races <- df[men_only,]
mens_races <- mens_races[order(mens_races[,1], mens_races[,7]),]
summary(mens_races)
```

Finally, it's time for a visual!
<br>
<br>
```{r echo=FALSE}
count_m <- mens_races[,c(1,5,6,7)]
count_f <- womens_races[,c(1,5,6,7)]
theme_set(theme_classic())
line_graph <- ggplot(count_m, aes(x=marathon, y=time))
line_graph + geom_dotplot(method="histodot", binaxis="y", binwidth=100, dotsize=0.75, stackdir="center", color="#00BFC4", fill="#00BFC4") +
  geom_dotplot(data=count_f, method="histodot", binaxis="y", binwidth=100, dotsize=0.75, stackdir="center", color="#F8766D", fill="#F8766D") +
  labs(title="Major Marathons, 1897-2018",
       subtitle="Winner finish time for men (teal) and women (salmon)",
       x="Marathon Name",
       y="Winner's Time (in seconds)")
```

We already knew that we had way more data for Boston, but the historical overlap in male v. female times is interesting. Let's look at a density chart for another perspective:
<br>
<br>
```{r echo=FALSE}
density <- df[,c(1,3,5,6,7)]
theme_set(theme_classic())
density_graph <- ggplot(density, aes(time))
density_graph + geom_density(aes(fill=factor(marathon)), alpha=0.8, show.legend=FALSE) +
  labs(title="Major Marathon Winning Times, 1897-2018",
       subtitle="Winner Finish Time Density, by race",
       x="Winner's Time (in seconds)",
       y="Density") +
  facet_grid(marathon ~ gender)
```

Men's finish times at Boston span a wide and relatively even range. Let's look at them by year, side-by-side with the women's finishes.
```{r echo=FALSE, fig.height=3}
boston_m <- mens_races[mens_races$marathon=="Boston",]
boston_f <- womens_races[womens_races$marathon=="Boston",]
boston_line <- ggplot() +
  geom_line(data=boston_m, aes(x=year, y=time), color="#00BFC4") +
  geom_line(data=boston_f, aes(x=year, y=time), color="#F8766D") +
  transition_reveal(year)
boston_line
```
Wow, the women were killing it in the 70s! :)

Now, suspend disbelief with me, if you will, and imagine that this data meets randomization requirements for standard statistical tests. Everyone seems to be getting faster, but the women are doing so at a greater velocity. I wonder if women were as fast in the past 20 years of racing to Boston's Copley Square as their male counterparts were 100 years earlier? Let's start with two quick visuals:
```{r fig.show="hold", out.width="50%", echo=FALSE}
men_boston_1899_1918 <- mens_races[mens_races$marathon=="Boston" & mens_races$year>=1899 & mens_races$year<=1918,]
women_boston_1999_2018 <- womens_races[womens_races$marathon=="Boston" & womens_races$year>=1999 & womens_races$year<=2018,]
plot(men_boston_1899_1918$year, men_boston_1899_1918$time,
  ylim=c(8400,10500),
  main="Boston Men's Wins, 1899-1918",
  xlab="Year",
  ylab="Winner's Finish Time (in seconds)")
plot(women_boston_1999_2018$year, women_boston_1999_2018$time,
  ylim=c(8400,10500),
  main="Boston Women's Wins, 1999-2018",
  xlab="Year",
  ylab="Winner's Finish Time (in seconds)")
```

Hmm, hard to tell from those scatterplots...maybe a boxplot will help?
<br>
<br>
```{r echo=FALSE}
men_boston_1899_1918 <- mens_races[mens_races$marathon=="Boston" & mens_races$year>=1899 & mens_races$year<=1918,]
women_boston_1999_2018 <- womens_races[womens_races$marathon=="Boston" & womens_races$year>=1999 & womens_races$year<=2018,]
boston_box <- ggplot(women_boston_1999_2018, aes(x=gender, y=time)) +
  geom_boxplot(fill="#F8766D", alpha=0.5) +
  geom_boxplot(data=men_boston_1899_1918, aes(x=gender, y=time), fill="#00BFC4", alpha=0.5) +
  scale_x_discrete(labels=c("Female" = "Women's\nFinishes\n1999-2018\nn=21", "Male" = "Men's\nFinishes\n1899-1918\nn=19")) +
  labs(title="Comparison of Boston Men v. Women",
       subtitle= "100 Years Apart",
       x="",
       y="Winner's Finish Time (in seconds)")
boston_box
```


It does look like the women of 1999-2018 were faster than the men of 1899-1918, but let's confirm that with an easy test.
```{r echo=FALSE, message=FALSE, warning=FALSE}
wilcox <- wilcox.test(women_boston_1999_2018$time, men_boston_1899_1918$time)
wilcox
```
Not a terribly useful bit of information that we've established here given that we're on shaky statistical ground, but pretty cool fun fact to drop at a cocktail party?

What else can we explore? The Kenyans are known for dominating the international marathoning world, right?
```{r echo=FALSE}
country_info <- as.factor(df$country)
summary(country_info)
```

Wow! 136 wins in this dataset! Let's zoom in on this...
<br>
<br>
```{r echo=FALSE}
kenyan_men <- mens_races[mens_races$country=="Kenya" & mens_races$year>=1983,]
kenyan_women <- womens_races[womens_races$country=="Kenya" & womens_races$year>=1983,]
kenyan_racers <- rbind(kenyan_men, kenyan_women)
theme_set(theme_gray())
kenyan_boxplot <- ggplot(kenyan_racers, aes(x=marathon, y=time, fill=gender))
kenyan_boxplot + geom_boxplot() +
  labs(title="Kenyan Major Marathon Wins, 1983-2018",
       subtitle="Winner Finish Times, by race and gender",
       x="Marathon Name",
       y="Winner's Finish Time (in seconds)")
```

It looks like the Kenyan men have high variability in finish times across races, but the women are pretty consistent; let's confirm with a Kruskal-Wallis test.
```{r echo=FALSE}
n_races_kenyan_men <- length(kenyan_men$year)
n_races_kenyan_women <- length(kenyan_women$year)
kruskal_men <- kruskal.test(time~marathon, kenyan_men)
kruskal_women <- kruskal.test(time~marathon, kenyan_women)
kruskal_results <- matrix(c(n_races_kenyan_men, kruskal_men$statistic, kruskal_men$parameter, round(kruskal_men$p.value,6), n_races_kenyan_women, kruskal_women$statistic, kruskal_women$parameter, round(kruskal_women$p.value,6)), ncol=4, byrow=TRUE)
colnames(kruskal_results) <- c("no_of_races", "chi_squared", "df", "p_value")
rownames(kruskal_results) <- c("kenyan_men", "kenyan_women")
kruskal_results
```
Nope, still plenty of variability in the Kenyan women across these races!  
<br>
<br>
Are you still suspending disbelief with me? Great, because it's time for a ridiculous ANOVA! Close your eyes while I check some standard assumptions, even though we're ignoring statistical reliability here...

##### Checking number of observations per group:
```{r echo=FALSE}
a <- length(kenyan_men$time[kenyan_men$marathon=="Boston"])
b <- length(kenyan_men$time[kenyan_men$marathon=="NYC"])
c <- length(kenyan_men$time[kenyan_men$marathon=="Chicago"])
d <- length(kenyan_men$time[kenyan_men$marathon=="London"])
e <- length(kenyan_men$time[kenyan_men$marathon=="Berlin"])
f <- length(kenyan_men$time[kenyan_men$marathon=="Tokyo"])
g <- length(kenyan_women$time[kenyan_women$marathon=="Boston"])
h <- length(kenyan_women$time[kenyan_women$marathon=="NYC"])
i <- length(kenyan_women$time[kenyan_women$marathon=="Chicago"])
j <- length(kenyan_women$time[kenyan_women$marathon=="London"])
k <- length(kenyan_women$time[kenyan_women$marathon=="Berlin"])
l <- length(kenyan_women$time[kenyan_women$marathon=="Tokyo"])
no_of_observations <- matrix(c(a,b,c,d,e,f,g,h,i,j,k,l), ncol=6, byrow=TRUE)
colnames(no_of_observations) <- c("Boston", "NYC", "Chicago", "London", "Berlin", "Tokyo")
rownames(no_of_observations) <- c("kenyan_men", "kenyan_women")
no_of_observations
```

##### Checking normality per group with Shapiro p-values:
(I'm zeroing out Tokyo wins for Kenyan women since there were only two observations.)
```{r echo=FALSE}
a <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="Boston"])$p.value
b <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="NYC"])$p.value
c <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="Chicago"])$p.value
d <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="London"])$p.value
e <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="Berlin"])$p.value
f <- shapiro.test(kenyan_men$time[kenyan_men$marathon=="Tokyo"])$p.value
g <- shapiro.test(kenyan_women$time[kenyan_women$marathon=="Boston"])$p.value
h <- shapiro.test(kenyan_women$time[kenyan_women$marathon=="NYC"])$p.value
i <- shapiro.test(kenyan_women$time[kenyan_women$marathon=="Chicago"])$p.value
j <- shapiro.test(kenyan_women$time[kenyan_women$marathon=="London"])$p.value
k <- shapiro.test(kenyan_women$time[kenyan_women$marathon=="Berlin"])$p.value
l <- 0   #not running shapiro for women's Tokyo since <3 observations and we're really just making stuff up at this point...
temp <- c(a,b,c,d,e,f,g,h,i,j,k)
temp <- round(temp, 6)
temp <- c(temp, l)
shapiro_p_values <- matrix(c(temp), ncol=6, byrow=TRUE)
colnames(shapiro_p_values) <- c("Boston", "NYC", "Chicago", "London", "Berlin", "Tokyo")
rownames(shapiro_p_values) <- c("kenyan_men", "kenyan_women")
shapiro_p_values
```

##### Checking homogeneity of variance per group:
```{r echo=FALSE}
a <- sd(kenyan_men$time[kenyan_men$marathon=="Boston"])
b <- sd(kenyan_men$time[kenyan_men$marathon=="NYC"])
c <- sd(kenyan_men$time[kenyan_men$marathon=="Chicago"])
d <- sd(kenyan_men$time[kenyan_men$marathon=="London"])
e <- sd(kenyan_men$time[kenyan_men$marathon=="Berlin"])
f <- sd(kenyan_men$time[kenyan_men$marathon=="Tokyo"])
g <- sd(kenyan_women$time[kenyan_women$marathon=="Boston"])
h <- sd(kenyan_women$time[kenyan_women$marathon=="NYC"])
i <- sd(kenyan_women$time[kenyan_women$marathon=="Chicago"])
j <- sd(kenyan_women$time[kenyan_women$marathon=="London"])
k <- sd(kenyan_women$time[kenyan_women$marathon=="Berlin"])
l <- sd(kenyan_women$time[kenyan_women$marathon=="Tokyo"])
sd_per_group <- matrix(c(a,b,c,d,e,f,g,h,i,j,k,l), ncol=6, byrow=TRUE)
colnames(sd_per_group) <- c("Boston", "NYC", "Chicago", "London", "Berlin", "Tokyo")
rownames(sd_per_group) <- c("kenyan_men", "kenyan_women")
sd_per_group
```

Ok, we're really far out on a precarious limb here and absolutely shouldn't be running any more statistical tests today since none of the right assumptions are in place...but we have to see this crazy ANOVA to its ridiculous conclusion!
```{r echo=FALSE}
summary(aov(time~marathon+gender+marathon:gender, kenyan_racers))
```

Unsurprisingly, we have main effects for gender and host city, but no interaction effect between those two variables. (I mean, if we hadn't thrown basic stats out the window hundreds of lines of code ago...) Here is a final visual to wrap things up!
<br>
<br>
```{r echo=FALSE}
a <- mean(kenyan_men$time[kenyan_men$marathon=="Boston"])
b <- mean(kenyan_men$time[kenyan_men$marathon=="NYC"])
c <- mean(kenyan_men$time[kenyan_men$marathon=="Chicago"])
d <- mean(kenyan_men$time[kenyan_men$marathon=="London"])
e <- mean(kenyan_men$time[kenyan_men$marathon=="Berlin"])
f <- mean(kenyan_men$time[kenyan_men$marathon=="Tokyo"])
g <- mean(kenyan_women$time[kenyan_women$marathon=="Boston"])
h <- mean(kenyan_women$time[kenyan_women$marathon=="NYC"])
i <- mean(kenyan_women$time[kenyan_women$marathon=="Chicago"])
j <- mean(kenyan_women$time[kenyan_women$marathon=="London"])
k <- mean(kenyan_women$time[kenyan_women$marathon=="Berlin"])
l <- mean(kenyan_women$time[kenyan_women$marathon=="Tokyo"])
marathon_names <- c("Boston", "NYC", "Chicago", "London", "Berlin", "Tokyo")
mens_times <- c(a,b,c,d,e,f)
womens_times <- c(g,h,i,j,k,l)
kenyan_avg <- data.frame(marathon_names, mens_times, womens_times)
colnames(kenyan_avg) <- c("marathon", "mean_finish_men", "mean_finish_women")
kenyan_dot <- ggplot(kenyan_avg, aes(x=marathon, y=mean_finish_men))
kenyan_dot + geom_point(color="#00BFC4", size=4) + geom_point(data=kenyan_avg, aes(x=marathon, y=mean_finish_women), color="#F8766D", size=4) +
  labs(title="Kenyan Marathon Wins, 1983-2018",
       subtitle="Average Winning Finish Times, by race and gender",
       x="Marathon Name",
       y="Mean Winning Time (in seconds)") +
theme(legend.position="none")
```